Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-11-13 main.tex
Addressed the review comment about background and connection of literature to our fall perturbations.

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage of reported serious injuries~\citep{Wegman2024}. At low speeds, from
start up to typical cruising speeds, bicycles are not self-stable and can be
challenging for the rider to balance. Low speed crashes may be reduced if the
bicycle was self-stable at these speeds by relieving the rider of some of their
necessary control activity. Bicycles can be mechanically modified to lower the
speeds at which they are self-stable~\citep{Astrom2005} and since the 1980s it
has been known that applying a motor actuated steering torque proportional to
the vehicle's roll angular rate~\footnote{``Roll'' is defined herein as per the
SAE J670 vehicle standards.} can stabilize a single-track vehicle down to very
low speeds~\citep{Ruijs1986}. If automatic control of steering can stabilize a
bicycle, it may reduce the control required from the rider to successfully
manage balancing tasks much like natural high speed self-stability already
does. We have developed a balance-assisting bicycle,
Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \citet{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall. From a mechanics perspective, it is not particularly important how or
where the perturbation is applied but that it results in forcing the
bicycle-rider system into an unrecoverable state, as all forces acting on a
dynamical system can be resolved into an equivalent force and torque of a
couple at any location in the system~\citep{Kane1985}. We chose to perturb the
handlebar because of this principle and it was practical and required a low
force to cause a fall.

>>>>

\section{Introduction}
%
Single-actor bicycle crashes are associated with a surprisingly large
percentage (\textgreater80\%) of reported serious injuries in The Netherlands
and values range betwen 50\% and 95\% in other surveyed
countries~\citep{Wegman2024}. \citet{Utriainen2023} report that slippery
surfaces, simple loss of control, and engagement with a curb, pothole, or
bollard are most the common fall types in the seven reported countries. Most of
these types involve perturbations that result in loss of balance. At low
speeds, from start up to typical cruising speeds, bicycles are not self-stable
and can be challenging for the rider to balance. Low speed crashes may be
reduced if the bicycle was self-stable at these speeds by relieving the rider
of some of their necessary control activity. Bicycles can be mechanically
modified to lower the speeds at which they are self-stable~\citep{Astrom2005}
and since the 1980s it has been known that applying a motor actuated steering
torque proportional to the vehicle's roll angular rate~\footnote{``Roll'' is
defined herein as per the SAE J670 vehicle standards.} can stabilize a
single-track vehicle down to very low speeds~\citep{Ruijs1986}. If automatic
control of steering can stabilize a bicycle, it may reduce the control required
from the rider to successfully manage balancing tasks much like natural high
speed self-stability already does. We have developed a balance-assisting
bicycle, Figure~\ref{fig:balance-assist-bicycle}, based on these principles and
hypothesize that it helps the rider in situations in which they are likely to
fall.
%
\begin{figure}
  \centering
  \includegraphics[width=80mm]{figures/balance-assist-bicycle.jpg}
  \caption{Balance assist bicycle prototype with electric motor in the steering
    column and data acquisition and control electronics mounted in the rear
    rack first introduced in \citet{Alizadehsaravi2023}.}
  \label{fig:balance-assist-bicycle}
\end{figure}

Riders fall when their control authority is not able to maintain a stable
vehicle state. There are many real world scenarios that can put the vehicle
into an uncontrollable state. External forces applied to the vehicle or rider
are one such scenario type and natural examples include wind gusts, handlebars
colliding with a neighbor's, a bag swinging from the handlebar, or simply
hitting a bump in the road. To assess our balance assisting bicycle, we subject
the rider to perturbations at the handlebar, which can easily cause a rider to
fall. From a mechanics perspective, it is not particularly important how or
where the perturbation is applied but that it results in forcing the
bicycle-rider system into an unrecoverable state, as all forces acting on a
dynamical system can be resolved into an equivalent force and torque of a
couple at any location in the system~\citep{Kane1985}. We chose to perturb the
handlebar because of this superposition principle and that it was both
practical and required a low force to cause a fall.
